---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# mut2

The aim om `mut2` is to provide functions for mutation models and exact likelihoods for pairs of individuals. Current functions include

* `lik2X()` : Pairwise likelihood for the X-chromosome
* `lik2()` : Pairwise likelihood for the X-chromosome
* `ibd1.parental` : Estimates for a pair of non-inbred individuals the probabilities of paternal origin when IBD is 1
* `makeReversible` : An irreducible and aperiodic mutation matrix
is tranformed to satisfy detailed balanced (DB) using the Metropolis-Hastings algorithm.

## Installation

To get the lastest version, install from GitHub as follows:

```{r, eval = FALSE}
 # First install devtools if needed
if(!require(devtools)) install.packages("devtools")

# Install mut2 from GitHub
devtools::install_github("thoree/mut2")
```

