---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(pedmut)
library(mut2, quietly = T)
```

# Example 1
Consider a marker with $n = 4$ alleles with allele frequencies
$p = (0.1, 0.3, 0.4, 0.2)$. Assume an equal mutation model with mutation rate $0.003$, which in this coincides with the expected mutation rate, i.e., 
\begin{align*}
Q=
		\begin{pmatrix}
			0.99700 & 0.00100 & 0.00100 & 0.00100 \\ 
			0.00100 & 0.99700 & 0.00100 & 0.00100 \\ 
			0.00100 & 0.00100 & 0.99700 & 0.00100 \\ 
			0.00100 & 0.00100 & 0.00100 & 0.99700 \\ 
		\end{pmatrix}
		.
\end{align*}
The  proposed mutation matrix and the allele frequency vector,
$(Q,p)$, does not correspond to a balanced mutation model. For instance
\[
p_2 q_{21} = 0.3\cdot 0.001 = 0.0003 \neq p_1q_{12} = 0.1\cdot 0.001 = 0.0001. 
\]
In this case the proposed matrix $Q$ is symmetric and
therefore transformation to reversibiliy with 'MH' option simplifies to 
\begin{align*}
		p_{ij}=q_{ij}\min \{1, \frac{p_j}{p_i} \}, \;i \neq j;
		p_{ii}=1-\sum_{i\neq j} p_{ij},
\end{align*}
Balancing is achieved by
```{r}
    Q = mutationMatrix("equal", rate = 0.003, alleles = 1:4)
    p = c(0.1,0.3,0.4,0.2)
    names(p) = 1:4
    P1 = findReversible(Q, afreq = p, method = 'MH')
    P1
```
The pair $(P_1,p)$ is balanced with expected mutation rate 0.002, which is lower than the expected for Q, 0.003. 	In this case the condition
$q_{ji} < p_i,\, j\neq i$ holds and we can transform using the alternative transformation (`method = PR`)
\begin{align}
p_{ij}=\frac{p_iq_{ij}+p_j q_{ji}}{2p_i},\;i \neq j,\;
p_{ii}=1-\sum_{i\neq j} p_{ij},
\end{align}
which preserves the expected mutation rate:
```{r}
    P2 = findReversible(Q, afreq = p, method = "PR")
    P2
```

Alternatively, the first balanced mutation matrix can be adjusted to have the required 
expected mutation rate

```{r}
    P3 = adjustReversible(Q, P1,  afreq = p)
    P3
```


