---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(pedmut, quietly = T)
library(pedsuite, quietly = T)
library(mut2, quietly = T)
```

# Example 1: Tutorial examples

Consider a marker with $n = 4$ alleles with allele frequencies
$p = (0.1, 0.3, 0.4, 0.2)$. Assume an equal mutation model with mutation rate $0.003$, which in this coincides with the expected mutation rate, i.e., 
\begin{align*}
Q=
		\begin{pmatrix}
			0.99700 & 0.00100 & 0.00100 & 0.00100 \\ 
			0.00100 & 0.99700 & 0.00100 & 0.00100 \\ 
			0.00100 & 0.00100 & 0.99700 & 0.00100 \\ 
			0.00100 & 0.00100 & 0.00100 & 0.99700 \\ 
		\end{pmatrix}
		.
\end{align*}
The  proposed mutation matrix and the allele frequency vector,
$(Q,p)$, does not correspond to a balanced mutation model. For instance
\[
p_2 q_{21} = 0.3\cdot 0.001 = 0.0003 \neq p_1q_{12} = 0.1\cdot 0.001 = 0.0001. 
\]
In this case the proposed matrix $Q$ is symmetric and
therefore default balancing simplifies to 
\begin{align*}
		p_{ij}=q_{ij}\min \{1, \frac{p_j}{p_i} \}, \;i \neq j;
		p_{ii}=1-\sum_{i\neq j} p_{ij},
\end{align*}
Balancing is achieved by
```{r}
    Q = mutationMatrix("equal", rate = 0.003, alleles = 1:4)
    p = c(0.1,0.3,0.4,0.2)
    names(p) = 1:4
    P1 = findReversible(Q, afreq = p)
    P1
```
The pair $(P_1,p)$ is balanced with expected mutation rate 0.002, which is lower than the expected for Q, 0.003. 	In this case the condition
$q_{ji} < p_i,\, j\neq i$ holds and we can transform using the alternative transformation (`method = PM`)
\begin{align}
p_{ij}=\frac{p_iq_{ij}+p_j q_{ji}}{2p_i},\;i \neq j,\;
p_{ii}=1-\sum_{i\neq j} p_{ij},
\end{align}
which preserves the expected mutation rate:
```{r}
    P2 = findReversible(Q, afreq = p, method = "PM")
    P2
```

Alternatively, the first balanced mutation matrix can be adjusted to have the required 
expected mutation rate

```{r}
    P1.adj = adjustReversible(Q, P1,  afreq = p)
    P1.adj
```
 
 There is also third option for balancing, the `Barker` option
```{r}
    P3 = findReversible(Q, afreq = p, method = "BA")
    P3
```

# Example 2: Extreme cases

## Trivial mutation matrix

Assume zero mutation rate. Then balancing leaves the matrix unchanged as it should:

```{r}
    Q = mutationMatrix("trivial", alleles = 1:4)
    p = c(0.1,0.3,0.4,0.2)
    names(p) = 1:4
    P= findReversible(Q, afreq = p)
    all(diag(P) == 1)
    P = findReversible(Q, afreq = p, method = "PM")
    all(diag(P) == 1)
```

## Always mutation

Consider 
```{r}
    Q = mutationMatrix("custom", matrix = matrix(c(0,1,1,0), ncol = 2),
                       alleles = 1:2)
    p = c(0.1,0.9)
    P = findReversible(Q, afreq = p)
    P
```
The two other methods, `PM` and `BA` do not work in this case.

## Equally likely mutations

In this case only "MH" and "BA" work
```{r}
    p = c(0.1,0.9)
    Q = mutationMatrix("custom", 
                       matrix = matrix(rep(0.5,4), ncol = 2),
                       alleles = 1:2, afreq = p)
    attr(Q, "rate") = expectedMutationRate(Q, p)
    P1 = findReversible(Q, afreq = p)
    P3 = findReversible(Q, afreq = p, method = "BA")
    cat("Expected mutation rates:", "\n")
    c("R.Q" = attr(Q, "rate"), "R.MH" = attr(P1, "rate"), 
      "R.BA" = attr(P3, "rate") )
```

## One step model

```{r}
    Q = mutationMatrix("onestep", rate = 0.003, alleles = 1:4)
    p = c(0.1,0.3,0.4,0.2)
    names(p) = 1:4
    P1 = findReversible(Q, afreq = p)
    P2 = findReversible(Q, afreq = p, method = "BA")
```
 
# SNPS

The input of the below function specifies
the allele frequency $p = (p_1, p_2)$ and a mutation matrix

\begin{align*}
Q =
	\begin{pmatrix}
		1-q_{12} & q_{12}  \\
		q_{21} & 1-q_{21}
	\end{pmatrix}
\end{align*}

LR.Q, the LR comparing "AF father of CH" to "AF and CH are unrelated", is
calculated for all genotype combinations.
The LR calculation is next done with Q replaced by a reversible
mutation matrix with the same expected mutation rate giving LR.M.
The ratio LRR = LR.M/LR.Q is found and also its expected value, ELRR.

```{r}
LRR = function(p1  = 0.5, q12 = 0.05, q21 = 0.05){
  p = c("1" = p1, "2" = 1 - p1)
  q = matrix(c(1 - q12, q21, q12, 1 - q21), ncol = 2)
  Q = mutationMatrix("custom", matrix = q, afreq = p, alleles = 1:2)

  r = expectedMutationRate(Q, p)
  # Reversible mutation model A with same mutation rate:
  if (max(r/(2*p[1]), r/(2*p[2])) > 1)
    stop("Reversible, mutation preserved matrix does not exist")
  A = matrix(c(1-r/(2*p[1]), r/(2*p[2]), r/(2*p[1]), 1-r/(2*p[2])), ncol = 2)
  
  x1 = nuclearPed(father = "AF", mother = "MO", child = "CH")
  m = marker(x1, afreq = p)
  x1 = setMarkers(x1, m)
  numerator = oneMarkerDistribution(x1, partialmarker = 1, ids = c("AF", "MO"), 
                                    verbose = F)
  x1 = setMutationModel(x1, "custom", matrix = A)
  L1 = oneMarkerDistribution(x1, partialmarker = 1, ids = c("AF", "CH"), 
                             verbose = F)
  x2 = setMutationModel(x1, "custom", matrix = Q)
  L2 = oneMarkerDistribution(x2, partialmarker = 1, ids = c("AF", "CH"), 
                             verbose = F)
  ELRR = sum(L1^2/L2)

  list(ELRR = ELRR,
       tab = list(LRM = L1/numerator,
                  LRQ = L2/numerator, LRR = L1/L2))
}
```

The default corresponds to a stationary model and hence we get an expected LRR equal to 1:

```{r}
LRR()$ELRR
```

Next follows an example where the input matrix is far from stationary:
```{r}
LRR(p1 = 0.01, q12 = 0.001, q21 = 0.001)
```

